---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
suppressPackageStartupMessages({
  library(minitypes)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/README-"
)
```

# minitypes


![](https://img.shields.io/badge/lifecycle-alpha-orange.svg)

`minitypes` is a collection of type conversion tools for R atomic vectors.

This package is built upon R's type conversion tools and aims to:

* have a consistent interface for converting between types
* add handling for types not handled by R explicitly e.g. `uint8`


## Supported Types:

* raw byte
    * This is R's builtin `raw` byte format. 
* `chr` 
    * Ascii/utf8 representation. 
* 1 bit:
    * `logical` - R's built-in logical type e.g. `c(T, T, F)`
    * `bit` - compact bit representation (from the [bit package](https://cran.r-project.org/package=bit))
    * `bitstring` - character representaction of bits e.g. `110`
* 8 bit 
    * `uint8` - R's built-in integer type. Checked to be within range [0 - 255]
    * `int8` - R's built-in integer type. Checked to be within range [-128, 127]
    * `hex8` - 2 character hexadecimal representation of a `uint8`
* 16 bit
    * `uint16` - R's built-in integer type. Checked to be within range [0 - 2^16-1]
    * `int16` - R's built-in integer type. Checked to be within range [-2^15, 2^15-1]
    * `hex16` - 4 character hexadecimal representation of a `uint16`
* 24 bit
    * `uint24` - R's built-in integer type. Checked to be within range [0 - 2^24-1]
    * `hex24` - 6 character hexadecimal representation of a `uint24`
* 32 bit
    * `int32` - R's built-in integer type

## Installation

You can install minitypes from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("coolbutuseless/minitypes")
```


## Promotion Rules

* types representing single bits will be padded out to be 
    * a multiple of 8 bits when converting to `raw` 
    * a multiple of 32 bits when converting to `int32`
    * the side to `pad` and the `value` (T/F) to pad with can be specified
* types representing 8-bits can be promoted to 16-bit and 32-bit types as long as the correct 
  number of bytes exist e.g.
    * can convert 4 `uint8` values to `int32`
    * can NOT convert 3 `raw` values to `uint16`


## Examples - Functions to convert supported types to/from `raw`

```{r}
r <- as.raw(c(1, 2, 0, 255))

raw_to_uint8(r)

raw_to_uint16(r, endian = 'little')

int32_to_raw(1L, endian = 'big')

raw_to_bitstring(r)

logical_to_raw(c(T, F, F), pad = 'left')
```


## `convert_type()` - conversion between any two supported types

All supported types can be converted to/from the `raw` type.  Thus any two 
supported types can be converted between
each other by using `raw` as the intermediary.

However, for complex endian/bit-order conversions (or cases where `raw` intermediary 
would lose information about the conversion), it is probably safer to do it
manually as a two-step process.


```{r example}
convert_type("1111111100010011", src_type='bitstring', dst_type='uint8')

convert_type(c(1L, 2L, 3L), src_type = 'int32', dst_type = 'uint8', endian = 'big')
```



## Examples - Specialised conversion functions

`minitypes` includes some functions for direct conversion between two supported 
types.

This may be for efficiency reasons, but is often because going through an 
intermediate `raw` step would lose information needed for the complete conversion.

```{r}
bitstring_to_logical("100 011")

int32_to_bitstring(1L, endian = 'big')

int32_to_bitstring(1L, endian = 'little')

int32_to_bitstring(1L, endian = 'little', first_bit = 'lsb')
```



## Helper functions

* `raw_reverse_bits()` - reverse the bits in a raw byte
* `raw_swap_endian()` - swap the order of a group of bytes (`nbytes` at a time) within a raw vector



# Useless tricks - `NA_integer_` & `NA_character_` bit representation

`NA_integer_` and `NA_character_` are just specific bit patterns within each of 
those types.

```{r}
convert_type(NA_integer_, 'int32', 'bitstring', endian = 'big', first_bit = 'msb')

convert_type(NA_character_, 'chr', 'bitstring', endian = 'big', first_bit = 'msb')
```



## Worked Example - Parsing a 'jpeg' header

* Read in a JPEG file as raw bytes
* Manually parse the JFIF header (see [JFIF wikipedia page](https://en.wikipedia.org/wiki/JPEG_File_Interchange_Format))
* JPG data is stored big endian

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Everyone with R has this jpeg. Read it in as 'raw' bytes
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
rjpeg_filename <- system.file('img/Rlogo.jpg', package='jpeg')
rjpeg          <- readBin(rjpeg_filename, 'raw', n = file.size(rjpeg_filename))

head(rjpeg, 50)
```

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Subset out the header components and convert to the correct type
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
raw_to_hex16 (rjpeg[ 1: 2], endian = 'big')  # JPEG SOI Marker $FFD8 (hexadecimal)

raw_to_hex16 (rjpeg[ 3: 4], endian = 'big')  # JFIF-APP0 Marker $FFE0 (hexadecimal)

raw_to_uint16(rjpeg[ 5: 6], endian = 'big')  # Length of APP0 Segment (integer)

raw_to_chr   (rjpeg[ 7:11], endian = 'big')  # 'JFIF' (text)
```





## Worked Example - Parsing a colour table from a 'gif'

* Read in a GIF file as raw bytes
* Manually parse the GIF header (see e.g. [GIF format](http://www.onicos.com/staff/iz/formats/gif.html))
* GIF data is stored little endian
* In most GIFs there's a global colour table with 255 colours (defined as a block of 255 RGB triplets)

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# A scarlet tanager!
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
gif_filename <- "figures/bird.gif"
gif          <- readBin(gif_filename, 'raw', n = file.size(gif_filename))

head(gif, 50)
```

`r glue::glue("![]({gif_filename})")`


```{r gif-colour-table}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Subset out the header components and convert to the correct type
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
raw_to_chr(gif[ 1: 6], endian = 'big')  # "GIF89a" (text)

raw_to_uint16(gif[ 7: 8], endian = 'little')  # width (integer)

raw_to_uint16(gif[ 9:10], endian = 'little')  # height (integer)

# Is a global colour table present?
colour_table_info <- raw_to_uint8(gif[11], endian = 'little')  
if (bitwAnd(colour_table_info, 128L)) { print("GCT present") }

# read 255 RGB triplets has 6-char-hex (i.e. hex24) starting at position 14
colour_table <- raw_to_hex24(gif[seq_len(255*3) + 13L], endian = 'big')
colour_table <- sort(paste0('#', colour_table))

head(colour_table)

image(x = 1, y = 1:255, z = matrix(1:255, nrow = 1), col = colour_table, axes = FALSE, ann = FALSE)
```





